Competing Constructions for German Measure Noun Phrases: from Usage Data to Experiment

### 2017-12-19 Latex/article.tex
Beginnig revisison (R3).

\usepackage[firstpage]{draftwatermark}
\SetWatermarkFontSize{30pt}
\SetWatermarkHorCenter{300pt}
\SetWatermarkColor[rgb]{0.9,0.5,0.5}
\SetWatermarkText{\textsf{Submitted April 2017. Do not quote.}}

@@

For such probabilistic generalisations estimated on corpus data, there has always been an interest in correlating the findings with results from experimental work (for example, \citealp{ArppeJaervikivi2007,BresnanEa2007,BresnanFord2010,DivjakGries2008,DivjakEa2016,FordBresnan2013}).
This often called a \textit{validation} of the corpus-derived findings, but \citet[303]{Divjak2016a} rightly criticises this choice of words ``because it creates the impression that behavioral experimental data is inherently more valuable than textual data,'' citing \cite{TummersEa2005}, who state that a corpus is ``a sample of spontaneous language use that is (generally) realized by native speakers.''
However, as \citet[486--487]{Dabrowska2016} convincingly argues, this does not imply that we can in some way ``deduce mental representations from patterns of use,'' \ie\ from corpus data.%
\footnote{In its colloquial meaning, the verb \textit{deduce} seems appropriate here.
However, in terms of the scientific method, we are most likely talking about \textit{induction} rather than \textit{deduction}.}
It would be highly surprising if this were possible and the same holds for experimental methods.
Nobody assumes that we can inductively infer mental representations from experiments, which -- as opposed to corpus studies -- even allow for direct access to the cognitive agent and offer much better possibilities to control experimental conditions and nuisance variables.
Rather, under the standard approach, a theory of cognitive representation is pre-specified.
Then, predictions are derived from this theory \textit{before} the experiment or the corpus study is conducted to \textit{test} the theory.
Technical problems of statistical inference aside, a successful experiment corroborates the theory, and a failed experiment weakens the theory.
The simplest assumption would be that the same approach is adequate for corpus studies.

The present study is conducted within the general paradigm of cognitive corpus linguistics and includes a comparison of the corpus findings with results from two experiments.
As a theoretical framework, I assume a model of similarity-based classification in the form of Prototype Theory.
Protoype Theory has been used in alternation research in cognitively oriented corpus linguistics (see \citealp{DivjakArppe2013}, \citealp{Gries2003}).
It is assumed that the variables annotated in a corpus study on an alternation phenomenon define prototypes for the alternants.
The similarity of a given feature vector (in a concrete sentence where one of the alternants is used) influences to a large extent which alternant is chosen by speakers.
This entails that a variant of Prototype Theory with features is assumed \citep{Rosch1978}, instead of the monolithic prototypes of earlier versions.
Prototype Theory is well suited for modeling constructional choices but it is just one of several similarity-based theories of classification, the most prominent other framework being Exemplar Theory (\citealp{MedinSchaffer1978,Hintzman1986}; see \citealp{StormsEa2000} for a comparison of the theories in experimental settings).
Prototype Theory and Exemplar Theory model essentially the same types of effects and differ mainly in whether they assume higher-level abstractions as part of the cognitive representation (Prototype Theory) or try to do without them (Exemplar Theory).
I am sceptical that corpus analysis alone could ever decide which theory is more suitable, given that substantial doubt has been voiced whether even experimental methods are ultimately able to do so \citep{Barsalou1990}.%
\footnote{In \cite{DivjakArppe2013}, it was shown using corpus data how both models form a converging picture.}
Prototype Theory (with feature abstractions) is preferred here simply because it fits the established alternation modeling paradigm, which relies on features being weighted in statistical models.
In Section~\ref{sec:analyses}, a prototype-theoretical parlance is therefore adopted.

In the remainder of the paper, I introduce the alternation phenomenon (Section~\ref{sec:descriptive}) and suggest a theory-driven set of factors influencing the alternation (Section~\ref{sec:analyses}).
Then the corpus study is presented, including an appropriate statistical analysis (Section~\ref{sec:corpusstudies}).
Two experiments that correlate results from the corpus study with experimental results are then reported (Section~\ref{sec:experimental}) before I sum up the paper in Section~\ref{sec:conclusion}.

@@

I use the term \textit{measure noun phrase} (MNP) to refer a noun phrase (NP) in which a kind-denoting (count or mass) noun depends on another noun that specifies a quantity of the objects or the substance denoted by the kind-denoting noun.
I call the kind-denoting noun the \textit{kind noun} and the quantity-denoting noun the \textit{measure noun}.
For illustration purposes, in the English \textit{a glass of good wine}, \textit{glass} is the measure noun and \textit{wine} is the kind noun.
Measure nouns can be all sorts of nouns which denote a quantity (such as \textit{litre} or \textit{amount}) but also those denoting containers, collections, etc. (such as \textit{glass} or \textit{bucket}).
Like \citet[284]{Brems2003}, I consider nouns as measure nouns ``which, strictly speaking, do not designate a `measure', but display a more nebulous potential for quantification'' (also \citealp[530]{Koptjevskaja2001}, and \citealp[338]{Rutkowski2007}).

@@

    \multirow{2}{*}{pseudo-partitive genitive} 
                & \multirow{2}{*}{---}                                  & \PGCa                                                   & \PGCd                      \\
                &                                                       & (\ref{ex:intro:alternation1}) \textit{Glas guten Weins} & (\ref{ex:intro:pseudopartitive1}) \textit{Glas dieses Weins} \\
  \end{tabular}
  \caption{Distribution of the NAC and PGC constructions in different NP structures with examples and references to full example sentences}
  \label{tab:constructions}
\end{table}

The distribution of the case patterns in the NAC (case identity between the measure noun and the kind noun) and in the PGC (genitive on the kind noun, regardless of the case of the measure noun) depending on the structure of the kind NP is summarised in Table~\ref{tab:constructions}.
I call the narrow apposition construction with a bare kind noun the \NACb\ and the partitive genitive with a determiner in the kind NP the \PGCd.
For the alternants with an adjective but no determiner in the kind noun phrase I use the terms \NACa\ and \PGCa.
In principle, this paper is about the middle column of Table~\ref{tab:constructions}, \ie\ the syntactic configuration in which two different case patterns are acceptable.
However, in Section~\ref{sec:analyses}, the outer columns (\NACb\ and \PGCd) will still play a major role when the factors controlling the alternation are named.

@@

As for (\ref{it:intro:noifl}), some measure nouns have been grammaticalised in a way that they always appear in their non-inflected form.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent', which have no normal plural forms at all.%
\footnote{Plurals like \textit{Pfunde} and \textit{Prozente} have special meanings and have very restricted uses, mostly in idiomatic expressions and prefabs such as \textit{Pfunde verlieren} `lose pounds.'
They cannot be used in normal MNPs.}
%\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass', but with a twist.
%A non-inflected version \textit{zwei Glas roter Wein} `two glasses of wine' and an inflected version \textit{zwei Gläser roter Wein} co-exist.
%The non-inflected version denotes the quantity of wine fitting into two glasses, the inflected version denotes two glasses filled with wine.}
I treat these cases like other measure nouns because they enter into both the \NACa\ as in (\ref{ex:uninflectedmeasures:a}) and the \PGCa\ as in (\ref{ex:uninflectedmeasures:b}).
In Section~\ref{sec:analyses}, however, degrees of grammaticalisation as a factor influencing the alternation will be discussed prominently.

@@

These construction have very idiosyncratic properties.
The construction with \textit{voller} is discussed in \cite{Zeldes2018} and the construction with \textit{mit} is discussed in \cite{Bhatt1990}.
They are not semantically equivalent to the NAC and the PGC and they are only available for a small subset of measure nouns.
All of these constructions are used only with measure nouns denoting containers, where the whole NP always refers to the container and never to the quantity contained in it.
They are consequently incompatible with measure nouns denoting natural portions such as \textit{Schluck} `gulp' or \textit{Haufen} `heap' and strongly grammaticalised nouns such as \textit{Gramm} `gram'.
This disqualifies them as alternatives to the \NACa\ or \PGCa\ in most contexts and they are thus decidedly \textit{not} more ways of saying the same thing. 
It is therefore reasonable to focus on the two well-defined alternants.

This concludes the descriptive overview of the phenomenon.
I have demonstrated that there is an alternation between two measure noun constructions in a narrow syntactic configuration (kind NP with an adjective but without a determiner), and that the two constructions differ in the case of the kind noun (case agreement with the measure noun or genitive).
I turn to more theory-oriented discussion in the next section.

@@

This section briefly reviews existing analyses of the \PGCa\ vs.\ \NACa\ alternation and related issues.
I also develop my own analysis and the appropriate hypotheses for the empirical studies presented in Sections~\ref{sec:corpusstudies} and \ref{sec:experimental}.
I discuss four main aspects:
first, the different syntactic structures of the alternants \NACa\ and \PGCa\ in relation to their syntactic prototypes \NACb\ and \PGCd;
second, degrees of grammaticalisation of the measure noun associated with the two prototypes;
third, a semantic prototypicality effect of the degrees of grammaticalisation (a preference to occur with cardinals);
fourth, preferences in different registers.

@@

First, my analysis is based on the idea that in the relevant syntactic structure [N\Sub{1}~[A~N\Sub{2}]\Sub{NP\Sub{2}}]\Sub{NP\Sub{1}} (as instantiated in the \NACa\ and the \PGCa), the adjective is morpho-syntactically ambiguous between a determiner and a modifying adjective.
To show this, %the first question is which noun constitutes the head of the whole construction.
%This was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
%Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
%Furthermore, the internal structure of MNPs can only be understood if the strong\slash weak inflection patterns of adjectives are taken into account.
the strong\slash weak inflection patterns of adjectives needs to be taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners themselves.
This is called the \textit{strong} inflectional pattern.
For example, in the dative (governed here by the preposition \textit{mit}), the strong suffix \textit{-em} is on the determiner in (\ref{ex:starkschwach:a}) and the adjective bears the weak suffix \textit{-en}. 
In (\ref{ex:starkschwach:b}), the strong suffix \textit{-em} appears on the adjective because there is no determiner.

@@

This morpho-syntactic ambiguity means that the \NACa\ is in fact a \NACb\ in disguise, and the \PGCa\ is a \PGCd\ in disguise.
This should ideally be reflected in the following basic selection effect:
It is known from frameworks like Collostructional Analysis \citep{GriesStefanowitsch2004} that lemmas are attracted with different strengths by constructions.
If the \NACa\ is highly similar to the \NACb\ and the \PGCa\ is highly similar to the \PGCd, the probability with which individual noun lemmas appear in the alternating constructions should be predictable at least in part from the relative frequency with which the same lemmas are used in the non-alternating constructions.
Nouns which occur proportionally more often in the \PGCd\ should favour the \PGCa, and nouns which occur proportionally more often in the \NACb\ should favour the \NACa.
This basic attraction effect will be used in the corpus study Section~\ref{sec:corpusstudies}.

@@

In a probabilistic framework like Prototype Theory, the crucial question (beyond simple lemma preference effects) is, however, what controls speakers' decisions to use either alternant. 
In the remainder of this section, I argue that the NAC and the PGC are prototypes associated with different degrees of the grammaticalisation of the measure noun and related morpho-syntactic properties, as well as register effects.
The degree of similarity of a given instance to either of the two prototypes makes speakers chose the \NACa\ or the \PGCa.

@@

In (\ref{ex:determiners:nac}), the measure noun is modified by a cardinal \textit{drei} `three', and hence the \NACa\ is preferred.
In (\ref{ex:determiners:pgc}), the measure noun is modified by a non-cardinal determiner \textit{einige} `some', and the \PGCa\ is preferred.
Especially with abstract physical measure nouns (like \textit{centilitre}), exact numerical quantification is invited.
By hpothesis, however, this goes beyond a selection effect tied to measure lemmas and cardinal quantifiers are expected to co-occur relatively more often with the \NACa.
The statistical model presented in Section~\ref{sec:corpusstudies} will be specified in a way that it can detect such a preference.

Fourth and finally, it was found that the \PGCa\ is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}).
This is not surprising, given that the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but by no means all) uses of the genitive are replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts and registers, such preferences can be part of what defines the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, register effects will therefore be modelled -- even if only using two very simple proxy variables.

To summarise the discussion and core hypotheses, I assume that the alternation is controlled by the similarity of the chosen lemmas (including the degree of grammaticalisation of measure nouns), certain morpho-syntactic choices, as well as the larger utterance context to two prototypes instantiated more straightforwardly by the two non-alternating cases of NAC and PGC.
Concretely:

\begin{enumerate}[i.]
  \item The relative frequencies with which measure noun lemmas and kind noun lemmas appear in the prototypical (non-alternating) \PGCd\ and \NACb\ are predictive of the probability with which the alternating \PGCa\ and the \NACa\ are chosen.
  \item (Classes of) more strongly grammaticalised measure nouns favour the \NACa.
  \item Measure nouns modified by cardinals favour the \NACa.
  \item The \PGCa\ is associated with higher registers.
\end{enumerate}

@@

In step (\ref{enum:data:step1}), I exported a list of all nouns in the DECOW14A01 sub-corpus sorted by their token frequency and manually went through it from the most frequent noun downwards, selecting the first one hundred mass nouns that occurred in the list.%
\footnote{DECOW14A01 is the first slice (roughly a twentieth) of the complete DECOW14A corpus.
It contains just over one billion tokens.}
Abstract nouns which partially behave like mass nouns (like \textit{Spaß} `fun’ or \textit{Gefahr} `danger’) were excluded because they are usually not quantified in the same way as concrete mass nouns.
The hundredth selected mass noun was \textit{Schmuck} `jewellery’, which is the 3,054th most frequent noun in the original frequency list.

@@

In step (\ref{enum:data:step3}), each of these 2,365 noun–noun combinations was queried in the target constructions (\PGCa\ and \NACa) individually in each of the first ten slices of DECOW (roughly 10 billion tokens).
In order to reduce the sample size for the manual annotation process, the final concordance was sampled from the results of these 2,365 queries.
Since the mass nouns in the sample were distributed according to the usual power law, I used all hits for nouns with a frequency up to 100 and a sample of 100 of all those with higher frequency.
The final sample contained 6,843 sentences, which was reduced to 5,063 in the manual annotation process due to removal of noisy material, erroneous hits and uninformative cases where the measure noun was in the genitive, in which case the \NACa\ cannot be distinguished from the \PGCa.
Given the careful sampling procedure described in this section, we can be highly sure that it contains all relevant and reasonably frequent noun–noun combinations in the target constructions.%
\footnote{In a similar fashion, the 100 most frequent measure nouns occurring with plural kind nouns were listed and queried, resulting in a sample of 871 sentences.
As stated in Section~\ref{sec:germanmeasurenps}, the \NACa\ is virtually never used with plural kind nouns, and this sample was not used except for quantifying the frequency of occurrence of the constructions (67 times \NACa\ and 794 times \PGCa).
However, the sample is distributed in the data package accompanying this paper.
}

@@

In Section~\ref{sec:analyses}, it was hypothesised that classes of measure lemmas might have different preferences for the two alternants.
To capture this, class information was annotated for measure lemmas.
The classification was inspired by the list in \citet[530]{Koptjevskaja2001} but due to the low frequencies of many of the potential classes, a very coarse classification was finally used.
With typical examples and their frequencies in the final sample, the classes are:
\textit{Physical} (abstract precisely measurable units such as \textit{Liter} `litre', \textit{Meter} `metre', \textit{Gramm} `gram'; \textit{f=}1,968),
\textit{Container} (\textit{Eimer} `bucket'; \textit{f=}740),
\textit{Amount} (\textit{Menge} `amount'; \textit{f=}1,364), 
\textit{Portion} (natural portions like \textit{Happen} `bite' or \textit{Krümel} `crumb'; \textit{f=}713).
The few lemmas that did not fit into either of these classes were labelled \textit{Rest} (\textit{f=}278).

@@

If there had been evidence in Levshina's study that Bayesian and MLE methods did \textit{not} converge, it would have been an occasion to demonstrate the selective superiority of the algorithms used in Bayesian estimation.
After all, there are situations where Bayesian estimators can be more robust, namely with heavily censored data, complex hierarchical models, perfect separation, etc. (see \citealp{Freedman1999}, \citealp[345--348]{GelmanHill2006}).
I want to state clearly that these points do not in any way invalidate the results presented in \cite{Levshina2016}.
However, being ``Bayesian'' is most likely not among its selling points.
Additionally, I want to voice the concern that many practitioners are probably already struggling with getting an adequate grasp of advanced statistical methods and that it might therefore be wise to use the more conservative and better understood method if the alternative method is not absolutely required for substantive reasons.
In Section~\ref{sec:corpushierarchicalmodel}, I estimate the parameters of my hierarchical model with MLE and Markov-Chain Monte Carlo (MCMC) methods (the currently most prominent estimator used in Bayesian settings) to demonstrate their expectable convergence.

>>>>

% \usepackage[firstpage]{draftwatermark}
% \SetWatermarkFontSize{30pt}
% \SetWatermarkHorCenter{300pt}
% \SetWatermarkColor[rgb]{0.9,0.5,0.5}
% \SetWatermarkText{\textsf{Submitted April 2017. Do not quote.}}

@@

For such probabilistic generalisations estimated on corpus data, there has always been an interest in correlating the findings with results from experimental work (for example, \citealp{ArppeJaervikivi2007,BresnanEa2007,BresnanFord2010,DivjakGries2008,DivjakEa2016,FordBresnan2013}).
This is often called a \textit{validation} of the corpus-derived findings, but \citet[303]{Divjak2016a} rightly criticises this choice of words ``because it creates the impression that behavioral experimental data is inherently more valuable than textual data,'' citing \cite{TummersEa2005}, who state that a corpus is ``a sample of spontaneous language use that is (generally) realized by native speakers.''
However, as \citet[486--487]{Dabrowska2016} convincingly argues, this does not imply that we can in some way ``deduce mental representations from patterns of use,'' \ie\ from corpus data.%
\footnote{In its colloquial meaning, the verb \textit{deduce} seems appropriate here.
However, in terms of the scientific method, we are most likely talking about \textit{induction} rather than \textit{deduction}.}
It would be highly surprising if this were possible and the same holds for experimental methods.
Nobody assumes that we can inductively infer mental representations from experiments, which -- as opposed to corpus studies -- even allow for direct access to the cognitive agent and offer much better possibilities to control experimental conditions and nuisance variables.
Rather, under the standard approach, a theory of cognitive representation is pre-specified.
Then, predictions are derived from this theory \textit{before} the experiment or the corpus study is conducted to \textit{test} the theory.
Technical problems of statistical inference aside, a successful experiment corroborates the theory, and a failed experiment weakens the theory.
The simplest assumption would be that the same approach is adequate for corpus studies.

The present study is conducted within the general paradigm of cognitive corpus linguistics and includes a comparison of the corpus findings with results from two experiments.
I assume a model of similarity-based classification in the form of Prototype Theory for modeling alternation.
Protoype Theory has been used in alternation research in cognitively oriented corpus linguistics (see \citealp{DivjakArppe2013}, \citealp{Gries2003}).
It is assumed that the variables annotated in a corpus study on an alternation phenomenon define prototypes for the alternants.
The similarity of a given feature vector (in a concrete sentence where one of the alternants is used) influences to a large extent which alternant is chosen by speakers.
This entails that a variant of Prototype Theory with features is assumed \citep{Rosch1978}, instead of the monolithic prototypes of earlier versions.
Prototype Theory is well suited for modeling constructional choices but it is just one of several similarity-based theories of classification, the most prominent other framework being Exemplar Theory (\citealp{MedinSchaffer1978,Hintzman1986}; see \citealp{StormsEa2000} for a comparison of the theories in experimental settings).
Prototype Theory and Exemplar Theory model essentially the same types of effects and differ mainly in whether they assume higher-level abstractions as part of the cognitive representation (Prototype Theory) or try to do without them (Exemplar Theory).
I am sceptical that corpus analysis alone could ever decide which theory is more suitable, given that substantial doubt has been voiced whether even experimental methods are ultimately able to do so \citep{Barsalou1990}.%
\footnote{In \cite{DivjakArppe2013}, it was shown using corpus data how both models form a converging picture.}
Prototype Theory (with feature abstractions) is preferred here simply because it fits the established alternation modeling paradigm, which relies on features being weighted in statistical models.
In Section~\ref{sec:analyses}, a prototype-theoretical parlance is therefore adopted.

In the remainder of the paper, I introduce the alternation phenomenon (Section~\ref{sec:descriptive}) and suggest a theory-driven set of factors influencing the alternation (Section~\ref{sec:analyses}).
Then the corpus study is presented, including an appropriate statistical analysis (Section~\ref{sec:corpusstudies}).
Two experiments which corroborate the corpus-based findings are then reported in Section~\ref{sec:experimental} before I sum up the paper in Section~\ref{sec:conclusion}.

@@

I use the term \textit{measure noun phrase} (MNP) to refer to a noun phrase (NP) in which a kind-denoting (count or mass) noun depends on another noun that specifies a quantity of the objects or the substance denoted by the kind-denoting noun.
I call the kind-denoting noun the \textit{kind noun} and the quantity-denoting noun the \textit{measure noun}.
For illustration purposes, in the English \textit{a glass of good wine}, \textit{glass} is the measure noun and \textit{wine} is the kind noun.
Measure nouns can be all sorts of nouns which denote a quantity (such as \textit{litre} or \textit{amount}) but also those denoting containers, collections, etc. (such as \textit{glass} or \textit{bucket}).
Like \citet[284]{Brems2003}, I consider nouns as measure nouns ``which, strictly speaking, do not designate a `measure', but display a more nebulous potential for quantification'' (also \citealp[530]{Koptjevskaja2001}, and \citealp[338]{Rutkowski2007}).

@@

    \multirow{2}{*}{pseudo-partitive genitive} 
                & \multirow{2}{*}{---}                                  & \PGCa                                                   & \PGCd                      \\
                &                                                       & (\ref{ex:intro:alternation1}) \textit{Glas guten Weins} & (\ref{ex:intro:pseudopartitive1}) \textit{Glas dieses Weins} \\
  \end{tabular}
  \caption{NAC and PGC constructions in different NP structures with examples and references to full example sentences}
  \label{tab:constructions}
\end{table}

The distribution of the case patterns in the NAC (case identity between the measure noun and the kind noun) and in the PGC (genitive on the kind noun, regardless of the case of the measure noun) depending on the structure of the kind NP is summarised in Table~\ref{tab:constructions}.
I call the narrow apposition construction with a bare kind noun the \NACb\ and the partitive genitive with a determiner in the kind NP the \PGCd.
For the alternants with an adjective but no determiner in the kind noun phrase I use the terms \NACa\ and \PGCa.
In principle, this paper is about the middle column of Table~\ref{tab:constructions}, \ie\ the syntactic configuration in which two different case patterns are acceptable.
However, in Section~\ref{sec:analyses}, the outer columns (\NACb\ and \PGCd) will still play a major role when the factors controlling the alternation are discussed.

@@

As for (\ref{it:intro:noifl}), some measure nouns have been grammaticalised in a way that they always appear in their non-inflected form.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent', which have no normal plural forms at all.%
\footnote{Plurals like \textit{Pfunde} and \textit{Prozente} have special meanings and have very restricted uses, mostly in idiomatic expressions and prefabs such as \textit{Pfunde verlieren} `lose pounds' and \textit{Prozente machen} `make a profit'.
They cannot be used in normal MNPs.}
%\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass', but with a twist.
%A non-inflected version \textit{zwei Glas roter Wein} `two glasses of wine' and an inflected version \textit{zwei Gläser roter Wein} co-exist.
%The non-inflected version denotes the quantity of wine fitting into two glasses, the inflected version denotes two glasses filled with wine.}
I treat these cases like other measure nouns because they enter into both the \NACa\ as in (\ref{ex:uninflectedmeasures:a}) and the \PGCa\ as in (\ref{ex:uninflectedmeasures:b}).
In Section~\ref{sec:analyses}, however, degrees of grammaticalisation as a factor influencing the alternation will be discussed prominently.

@@

These construction have very idiosyncratic properties.
The construction with \textit{voller} is discussed in \cite{Zeldes2018} and the construction with \textit{mit} is discussed in \cite{Bhatt1990}.
They are not semantically equivalent to the NAC and the PGC and they are only available for a small subset of measure nouns.
All of these constructions are used only with measure nouns denoting containers, where the whole NP always refers to the container and never to the quantity contained in it.
They are consequently incompatible with measure nouns denoting natural portions such as \textit{Schluck} `gulp' or \textit{Haufen} `heap' and strongly grammaticalised nouns such as \textit{Gramm} `gram'.
This disqualifies them as alternatives to the \NACa\ or \PGCa\ in most contexts and they are thus clearly not just more ways of saying the same thing. 
It is therefore reasonable to focus on the two well-defined alternants.

This concludes the descriptive overview of the phenomenon.
I have demonstrated that there is an alternation between two measure noun constructions in a narrow syntactic configuration (kind NP with an adjective but without a determiner), and that the two constructions differ in the case of the kind noun (case agreement with the measure noun or genitive).
I turn to more theory-oriented discussion of the alternation in the next section.

@@

This section briefly reviews existing analyses of the \PGCa\ vs.\ \NACa\ alternation and related issues.
I also develop my own analysis and the appropriate hypotheses for the empirical studies presented in Sections~\ref{sec:corpusstudies} and \ref{sec:experimental}.
I discuss four main aspects:
first, the different syntactic structures of the alternants \NACa\ and \PGCa\ in relation to their syntactic prototypes \NACb\ and \PGCd;
second, degrees of grammaticalisation of the measure noun associated with the two prototypes;
third, a semantic prototypicality effect of the degrees of grammaticalisation (a preference to occur with cardinals);
fourth, preferences in different styles.

@@

First, my analysis is based on the idea that in the relevant syntactic structure [N\Sub{1}~[A~N\Sub{2}]\Sub{NP\Sub{2}}]\Sub{NP\Sub{1}} (as instantiated in the \NACa\ and the \PGCa), the adjective is morpho-syntactically ambiguous between a determiner and a modifying adjective.
To show this, %the first question is which noun constitutes the head of the whole construction.
%This was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
%Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
%Furthermore, the internal structure of MNPs can only be understood if the strong\slash weak inflection patterns of adjectives are taken into account.
the strong\slash weak inflection patterns of adjectives needs to be taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners themselves.
This is called the \textit{strong} inflectional pattern.
For example, in the dative (governed here by the preposition \textit{mit}), the strong suffix \textit{-em} is on the determiner in (\ref{ex:starkschwach:a}) and the adjective bears the weak suffix \textit{-en}. 
In (\ref{ex:starkschwach:b}), the strong suffix \textit{-em} appears on the adjective because there is no determiner.%
\footnote{In the masculine and neuter singular genitive, the strong and weak forms are indistinguishable.
Since the alternation itself does not occur in the genitive (see Section~\ref{sec:descriptive}), this does not create any complications for the present study.}

@@

This morpho-syntactic ambiguity means that the \NACa\ is in fact a \NACb\ in disguise, and the \PGCa\ is a \PGCd\ in disguise.
This should ideally be reflected in the following basic selection effect:
It is known from frameworks like Collostructional Analysis \citep{GriesStefanowitsch2004} that lemmas are attracted with different strengths by constructions.
If the \NACa\ is highly similar to the \NACb\ and the \PGCa\ is highly similar to the \PGCd, the probability with which individual noun lemmas appear in the alternating constructions should be predictable at least in part from the relative frequency with which the same lemmas are used in the non-alternating constructions.
Nouns which occur proportionally more often in the \PGCd\ should favour the \PGCa, and nouns which occur proportionally more often in the \NACb\ should favour the \NACa.
This basic attraction effect will be quantified in the corpus study Section~\ref{sec:corpusstudies}.

@@

In a probabilistic framework like Prototype Theory, the crucial question (beyond simple lemma preference effects) is, however, which factors control speakers' decisions to use either alternant. 
In the remainder of this section, I argue that the NAC and the PGC are prototypes associated with different degrees of the grammaticalisation of the measure noun and related morpho-syntactic properties, as well as stylistic effects.
The degree of similarity of a given instance to either of the two prototypes makes speakers chose the \NACa\ or the \PGCa.

@@

In (\ref{ex:determiners:nac}), the measure noun is modified by a cardinal \textit{drei} `three', and hence the \NACa\ is preferred.
In (\ref{ex:determiners:pgc}), the measure noun is modified by a non-cardinal determiner \textit{einige} `some', and the \PGCa\ is preferred.
Especially with exact physical measure nouns (like \textit{centilitre}), exact numerical quantification is invited.
By hypothesis, however, this goes beyond a selection effect tied to measure lemmas and cardinal quantifiers are expected to co-occur relatively more often with the \NACa.
This hypothesis will be reflected in the specification of the statistical model presented in Section~\ref{sec:corpusstudies}.

Fourth and finally, it was found that the \PGCa\ is more typical of higher stylistic levels (distinctly edited, closer to the non-regional standard, more formal) and\slash or even exclusive to written language (see \citealp[320--323]{Hentschel1993}).
This is not surprising, given that the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but by no means all) uses of the genitive are replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts, registers, and styles, such preferences can be part of what defines the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, stylistic effects will therefore be modelled -- even if only using two very simple proxy variables.

To summarise, I assume that the alternation is controlled by the following factors:

\begin{enumerate}[i.]
  \item The relative frequencies with which measure noun lemmas and kind noun lemmas appear in the prototypical (non-alternating) \PGCd\ and \NACb\ are predictive of the probability with which the alternating \PGCa\ and the \NACa\ are chosen.
  \item (Classes of) more strongly grammaticalised measure nouns favour the \NACa.
  \item Measure nouns modified by cardinals favour the \NACa.
  \item The \PGCa\ is associated with higher stylistic levels.
\end{enumerate}

@@

In step (\ref{enum:data:step1}), I exported a list of all nouns in the DECOW14A01 sub-corpus sorted by their token frequency and manually went through it from the most frequent noun downwards, selecting the first one hundred mass nouns that occurred in the list.%
\footnote{DECOW14A01 is the first slice (roughly a twentieth) of the complete DECOW14A corpus.
It contains just over one billion tokens.}
Mass nouns were defined as concrete nouns which denote a substance in the broad sense, combine with uninflected mass quantifiers such as \textit{viel} `much' and \textit{wenig} `little' (\textit{viel Bier} `much beer'), and form only sortal and unit plurals (such as the plural \textit{Biere} `types of beer' or `glasses of beer').
Abstract nouns which partially behave like mass nouns (like \textit{Spaß} `fun’ or \textit{Gefahr} `danger’) were excluded because they are usually not quantified in the same way as concrete mass nouns.
The hundredth selected mass noun was \textit{Schmuck} `jewellery’, which is the 3,054th most frequent noun in the original frequency list.

@@

In step (\ref{enum:data:step3}), each of these 2,365 noun–noun combinations was queried in the target constructions (\PGCa\ and \NACa) individually in each of the first ten slices of DECOW (roughly 10 billion tokens).
In order to reduce the sample size for the manual annotation process, the final concordance was sampled from the results of these 2,365 queries.
Since the mass nouns in the sample were distributed according to the usual power law (often referred to as a \textit{Zipfian} distribution), I used all hits for nouns with a frequency up to 100 and a sample of 100 of all those with higher frequency.
The final sample contained 6,843 sentences, which was reduced to 5,063 in the manual annotation process due to removal of noisy material, erroneous hits and uninformative cases where the measure noun was in the genitive, in which case the \NACa\ cannot be distinguished from the \PGCa.
Given the careful sampling procedure described in this section, we can be highly sure that it contains all relevant and reasonably frequent noun–noun combinations in the target constructions.%
\footnote{In a similar fashion, the 100 most frequent measure nouns occurring with plural kind nouns were listed and queried, resulting in a sample of 871 sentences.
As stated in Section~\ref{sec:germanmeasurenps}, the \NACa\ is virtually never used with plural kind nouns, and this sample was not used except for quantifying the frequency of occurrence of the constructions (67 times \NACa\ and 794 times \PGCa).
However, the sample is distributed in the data package accompanying this paper.
}

@@

In Section~\ref{sec:analyses}, it was hypothesised that classes of measure lemmas might have different preferences for the two alternants.
To capture this, class information was annotated for measure lemmas.
The classification was inspired by the list in \citet[530]{Koptjevskaja2001} but due to the low frequencies of many of the potential classes, a very coarse classification was finally used.
With typical examples and their frequencies in the final sample, the classes are:
\textit{Physical} (abstract precisely measurable units such as \textit{Liter} `litre', \textit{Meter} `metre', \textit{Gramm} `gram'; \textit{f=}1,968),
\textit{Container} (\textit{Eimer} `bucket'; \textit{f=}740),
\textit{Amount} (\textit{Menge} `amount'; \textit{f=}1,364), 
\textit{Portion} (natural portions like \textit{Happen} `bite' or \textit{Krümel} `crumb'; \textit{f=}713).
The lemmas that did not fit into either of these classes were labelled \textit{Rest} (\textit{f=}278).

@@

If there had been evidence in Levshina's study that Bayesian and MLE methods did \textit{not} converge, it would have been an occasion to demonstrate the selective superiority of the algorithms used in Bayesian estimation.
After all, there are situations where Bayesian estimators can be more robust, namely with heavily censored data, complex hierarchical models, perfect separation, etc. (see \citealp{Freedman1999}, \citealp[345--348]{GelmanHill2006}).
I want to state clearly that these points do not in any way invalidate the results presented in \cite{Levshina2016}.
However, being ``Bayesian'' is most likely not among its selling points.
Additionally, I want to voice the concern that many practitioners are probably already struggling with getting an adequate grasp of advanced statistical methods and that it might therefore be wise to use the more conservative and better understood method if the alternative method is not absolutely required for substantive reasons.
In the next section, I estimate the parameters of my hierarchical model with MLE and Markov-Chain Monte Carlo (MCMC) methods (the currently most prominent estimator used in Bayesian settings) to demonstrate their expectable convergence.
